---
title: "DSTI - Survival Analysis Project"
author: "Samd Guizani"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Imports

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyr)
library(broom)
library(survival) # Survival analysis toolbox
library(gtsummary)
library(ggfortify)
library(ggsurvfit)
library(survminer) # For survival analysis plots 
library(gridExtra)
library(muhaz) # Hazard function estimation
```

# Functions

```{r}

```

# Main Script

## Data Loading & Preprocessing

-   `NewSpring.csv` is a publicly available dataset from *William Q. Meeker, Luis A. Escobar, Francis G. Pascual, **Statistical Methods for Reliability Data, 2nd Edition*** ([Statistical Methods for Reliability Data, 2nd Edition \| Wiley](https://www.wiley.com/en-us/Statistical+Methods+for+Reliability+Data%2C+2nd+Edition-p-9781118115459))
-   The file, among other datasets focusing on survival and reliability analysis, are available on *DataShare, Iowa State University’s open data repository* ([Item - Reliability Data: Laboratory Failure-Time and Single-Time Strength Data - Iowa State University - Figshare](https://iastate.figshare.com/articles/dataset/Reliability_Data_Laboratory_Failure-Time_and_Single-Time_Strength_Data/14454753?file=28383774))
-   The dataset records the ***number of cycles to failure*** for a spring under different conditions, including:
    -   ***Stroke length*** (factor A, numeric, 50 to 70 mils)
    -   ***Processing temperature*** (factor B, numeric 500 to 1000 F)
    -   ***Method used*** (factor C, categorical, New vs. Old)
-   The experiment is organized as a **Full Factorial Design**. For each combination of the 3 factors, 9 units are tested for a maximum of 5000 cycles. The `Censoring` column indicates whether the observation is censored (0, the spring survived 5000 cycles) or an event (1, number of cycles until failure is recorded)
-   The `Count` column represents the number of occurrences for each combination of conditions reaching an identical outcome.\
    *E.g.: a row where Censoring = 0 (Censored) and Count = 3 means 3 units reached the study end (5000 cycles) and removed from the test while still functioning.*

```{r}
# Import data from folder Datasets
df <- read.csv("./Datasets/NewSpring.csv")

# Rename columns for better readability
colnames(df) = c('Cycles', 'A.Stroke', 'B.ProcessTpt', 'C.Method', 'Censoring', 'Count')
factor_names = colnames(df)[2:4]

# Uncount the data by replicating rows according to the 'Count' column
df <- df |> uncount(Count)

# Convert Method to factor
df$C.Method <- as.factor(df$C.Method)

# Convert Censoring to factor (0 for censored, 1 for event)
df$Censoring <- ifelse(df$Censoring == "Failed", 1, 0)
```

## Exploratory Data Analysis

```{r}
for (f in factor_names) {
  boxplot(Cycles ~ get(f), 
          data = df, 
          main = paste("Boxplot of Cycles by", f), 
          xlab = f, 
          ylab = "Cycles to Failure")
}
```

```{r}
# Plot 'Cycles' vs. 'A.Stroke'
# Visualize other factors as color or facets
print(ggplot(df, aes(x = A.Stroke, y = Cycles, colour = as.factor(B.ProcessTpt))) +
        geom_point() +
        geom_smooth(method = "lm", se = TRUE) +
        facet_wrap(~ C.Method)
      )
```

## Survival Analysis

```{r}
# Fit Kaplan-Meier model for each factor

for (f in factor_names) {
  fit <- survfit(
    Surv(Cycles, Censoring) ~ get(f), 
    data = df,
    conf.type = "log-log")
  
  # Plot the curves
  ggsurvplot(fit, data = df, 
             conf.int = TRUE, 
             pval = TRUE, 
             title = paste("Survival Curves by", f)) |> print()
  
}
```

```{r}
# Compare groups using Logrank or Fleming-Harrington tests 
rho = 1 # 0 for Logrank, 1 for FH test
survdiff(
  Surv(Cycles, Censoring) ~ A.Stroke + B.ProcessTpt + C.Method, data = df, rho = rho)
```

```{r}
# Example for comparing two groups

# old_haz <- with(subset(df, A.Stroke == 50), kphaz.fit(Cycles, Censoring))
# new_haz <- with(subset(df, A.Stroke == 70), kphaz.fit(Cycles, Censoring))

# old_haz <- with(subset(df, B.ProcessTpt == 500), kphaz.fit(Cycles, Censoring))
# new_haz <- with(subset(df, B.ProcessTpt == 1000), kphaz.fit(Cycles, Censoring))

old_haz <- with(subset(df, C.Method == "Old"), kphaz.fit(Cycles, Censoring))
new_haz <- with(subset(df, C.Method == "New"), kphaz.fit(Cycles, Censoring))

plot(old_haz$time, old_haz$haz, type='l', col="red", 
     ylim = c(0, max(old_haz$haz, new_haz$haz))
     )
lines(new_haz$time, new_haz$haz, col="blue")
```

```{r}
# Model with main effects and the interaction between Stroke and Method
cox_interaction <- coxph(
  Surv(Cycles, Censoring) ~ A.Stroke + B.ProcessTpt + C.Method,
  data = df)
cox_interaction
```
