---
title: "DSTI - Survival Analysis Project"
author: "Samd Guizani"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Imports

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyr)
library(broom)
library(survival) # Survival analysis toolbox
library(gtsummary)
library(ggfortify)
library(ggsurvfit)
library(survminer) # For survival analysis plots 
library(gridExtra)
library(muhaz) # Hazard function estimation
library(plyr) # For data manipulation in survival analysis
```

# Functions

```{r}

```

# Main Script

## Data Loading & Preprocessing

-   `NewSpring.csv` is a publicly available dataset from *William Q. Meeker, Luis A. Escobar, Francis G. Pascual, **Statistical Methods for Reliability Data, 2nd Edition*** ([Statistical Methods for Reliability Data, 2nd Edition \| Wiley](https://www.wiley.com/en-us/Statistical+Methods+for+Reliability+Data%2C+2nd+Edition-p-9781118115459))
-   The file, among other datasets focusing on survival and reliability analysis, are available on *DataShare, Iowa State University’s open data repository* ([Item - Reliability Data: Laboratory Failure-Time and Single-Time Strength Data - Iowa State University - Figshare](https://iastate.figshare.com/articles/dataset/Reliability_Data_Laboratory_Failure-Time_and_Single-Time_Strength_Data/14454753?file=28383774))
-   The dataset records the ***number of cycles to failure*** for a spring under different conditions, including:
    -   ***Stroke length*** (factor A, numeric, 50 to 70 mils)
    -   ***Processing temperature*** (factor B, numeric 500 to 1000 F)
    -   ***Method used*** (factor C, categorical, New vs. Old)
-   The experiment is organized as a **Full Factorial Design**. For each combination of the 3 factors, 9 units are tested for a maximum of 5000 cycles. The `Censoring` column indicates whether the observation is censored (0, the spring survived 5000 cycles) or an event (1, number of cycles until failure is recorded)
-   The `Count` column represents the number of occurrences for each combination of conditions reaching an identical outcome.\
    *E.g.: a row where Censoring = 0 (Censored) and Count = 3 means 3 units reached the study end (5000 cycles) and removed from the test while still functioning.*

```{r}
# Import data from folder Datasets
df <- read.csv("./Datasets/NewSpring.csv")

# Rename columns for better readability
colnames(df) = c('Cycles', 'A.Stroke', 'B.ProcessTpt', 'C.Method', 'Censoring', 'Count')
factor_names = colnames(df)[2:4]

# Uncount the data by replicating rows according to the 'Count' column
df <- df |> uncount(Count)

# Convert Method to factor
df$C.Method <- as.factor(df$C.Method)
df$C.Method <- relevel(df$C.Method, ref = "Old" ) # "Old" considered as reference

# Convert Censoring to factor (0 for censored, 1 for event)
df$Censoring <- ifelse(df$Censoring == "Failed", 1, 0)
```

## Exploratory Data Analysis

### Boxplots

```{r}
par(mfrow = c(1, 3))
for (f in factor_names) {
  boxplot(Cycles ~ get(f), 
          data = df, 
          main = paste("Boxplot of Cycles by", f), 
          xlab = f, 
          ylab = "Cycles to Failure")
}
```

#### Observations:

-   The distribution of the number of cycles to spring failure graphically indicate a strong effect of covariate A and C.

-   Covariate B seems to have less influence on the failure rate.

### Linear "trending"

```{r}
# Plot 'Cycles' vs. 'A.Stroke'
# Visualize other factors as color or facets
print(ggplot(df, aes(x = A.Stroke, y = Cycles, colour = as.factor(B.ProcessTpt))) +
        geom_point() +
        geom_smooth(method = "lm", se = TRUE) +
        facet_wrap(~ C.Method)
      )
```

#### Observations:

-   ***ATTENTION***: Fitting a linear regression is only made for *informative purpose*. Censored data should not be modeled using linear regression models.

-   Increasing A -\> increasing failure rate

-   Increasing B -\> increasing failure rate

-   New method (C) -\> decreasing failure rate

## Survival Analysis

### Kaplan-Meir models [1 covariate at a time]

```{r}
# Fit Kaplan-Meier model for each covariate
for (f in factor_names) {
  fit <- survfit(
    Surv(Cycles, Censoring) ~ get(f), 
    data = df,
    conf.type = "log-log")
  
  print(paste0('-------- Survival model with covariate: ', f, ' --------'))
  cat('\n')
  
  print(fit)
  
  cat('\n')
  print('---------------------------------------------------------------')
  cat('\n')
  
  # Plot the curves
  ggsurvplot(fit, data = df, 
             conf.int = TRUE, 
             pval = TRUE, 
             title = paste("Survival Curves by", f)) |> print()
  
}
```

#### Observations:

-   A.Stroke and C.Method show significant differences (p \< 0.05) in survival curves

-   B.ProcessTpt shows less pronounced differences (p = 0.051)

### Logrank comparison tests [1 covariate at a time, with stratification]

```{r}
# Compare groups using Logrank or Fleming-Harrington tests 

rho = 0 # 0 for Logrank, 1 for FH test

for (f in factor_names) {
  # 1. Identify which variables to stratify by
  strata_vars <- factor_names[factor_names != f]
  
  # 2. Build the formula string
  # Result looks like: "Surv(Cycles, Censoring) ~ A.Stroke + strata(B.ProcessTpt, C.Method)"
  strata_part <- paste0("strata(", paste(strata_vars, collapse = ", "), ")")
  formula_str <- paste0("Surv(Cycles, Censoring) ~ ", f, " + ", strata_part)
  
  print(paste0('-------- Testing Covariate: ', f, ' --------'))
  print(paste0('Formula: ', formula_str))
  
  # 3. Convert string to formula and run test
  test_result <- survdiff(as.formula(formula_str), data = df, rho = rho)
  
  print(test_result)
  cat('\n---------------------------------------------------\n\n')
}
```

#### Observations:

-   A, B and C show statistically significant differences (p \< 0.05) in survival curves

### Hazard functions at low vs. high levels of each covariate

```{r}
par(mfrow = c(1, 3))

# Comparing A.Stroke = 50 vs. 70

haz_low <- with(subset(df, A.Stroke == 50), kphaz.fit(Cycles, Censoring))
haz_high <- with(subset(df, A.Stroke == 70), kphaz.fit(Cycles, Censoring))

plot(haz_low$time, haz_low$haz, type='l', col="red", 
     xlab = 'Cyles',
     ylab = 'h(t)',
     ylim = c(0, max(haz_low$haz, haz_high$haz))
     )
lines(haz_high$time, haz_high$haz, col="blue")

# Comparing B.ProcessTpt = 500 vs. 1000

haz_low <- with(subset(df, B.ProcessTpt == 500), kphaz.fit(Cycles, Censoring))
haz_high <- with(subset(df, B.ProcessTpt == 1000), kphaz.fit(Cycles, Censoring))

plot(haz_low$time, haz_low$haz, type='l', col="red", 
     xlab = 'Cyles',
     ylab = 'h(t)',
     ylim = c(0, max(haz_low$haz, haz_high$haz))
     )
lines(haz_high$time, haz_high$haz, col="blue")

# Comparing two groups C.Method = Old vs. New

haz_low <- with(subset(df, C.Method == "Old"), kphaz.fit(Cycles, Censoring))
haz_high <- with(subset(df, C.Method == "New"), kphaz.fit(Cycles, Censoring))

plot(haz_low$time, haz_low$haz, type='l', col="red", 
     xlab = 'Cyles',
     ylab = 'h(t)',
     ylim = c(0, max(haz_low$haz, haz_high$haz))
     )
lines(haz_high$time, haz_high$haz, col="blue")
```

#### Observations:

-   Graphs not easy to interpret, but can observe the following:

    -   A, hazard function for 70 (blue) higher than 50 (red) -\> higher failure rate over time.

    -   B, hazard function for 500 (blue) higher than 1000 (red) -\> higher failure rate over time.

    -   C, hazard function for "New" (blue) lower than "Old" (red) -\> lower failure rate over time.

### Cox Proportional Hazard (CPH) modelling

```{r}
# Model with main effects and the interaction between Stroke and Method
fit.cph_full <- coxph(
  Surv(Cycles, Censoring) ~ A.Stroke + B.ProcessTpt + C.Method + 
    A.Stroke * B.ProcessTpt + A.Stroke * C.Method + B.ProcessTpt * C.Method,
  data = df)
summary(fit.cph_full)
```

##### Observations:

-   Full model with all main effects and interactions shows that A.Stroke is significant (p \< 0.05)

-   Model needs to be simplified

### CPH Model selection with AIC selection using step() function

```{r}
# Perform model selection based on AIC
fit.cph_selected <- step(fit.cph_full, direction = "both", trace = 2)
```

```{r}
# Display the summary of the selected model
summary(fit.cph_selected)
```

#### Observations:

-   The selected model includes A, B and C which are all significant (p \< 0.05)
-   No interaction terms included in the final model, suggesting that the effects of the covariates are additive rather than multiplicative.
-   Results from CPH model after term selection are consistent with previous analysis (Kaplan-Meier, Logrank tests, hazard functions) indicating that A, B and C have significant effects on the survival of the springs.

### Checking CPH assumption: Martingale residuals

```{r}
# Martingale residuals for the selected model
martingale_residuals <- residuals(fit.cph_selected, type = "martingale")
df$Resid.Martingale <- martingale_residuals

# plot residuals vs factors
par(mfrow = c(1, 3))

with(df, {
  plot(A.Stroke, Resid.Martingale, 
       xlab = "A.Stroke", 
       ylab = "Martingale Residuals", 
       main = "Martingale Resid. vs. A.Stroke")
  lines(lowess(A.Stroke, Resid.Martingale), col = "red")
  
  plot(B.ProcessTpt, Resid.Martingale, 
       xlab = "B.ProcessTpt", 
       ylab = "Martingale Residuals", 
       main = "Martingale Resid. vs. B.ProcessTpt")
  lines(lowess(B.ProcessTpt, Resid.Martingale), col = "red")
  
  plot(C.Method, Resid.Martingale, 
       xlab = "C.Method", 
       ylab = "Martingale Residuals", 
       main = "Martingale Resid. vs. C.Method")
})

par(mfrow = c(1, 1))
# Plot Martingale residuals against fitted values
plot(fitted(fit.cph_selected), martingale_residuals,
     xlab = "Fitted Values",
     ylab = "Martingale Residuals",
     main = "Martingale Residuals vs. Fitted Values")
lines(lowess(fitted(fit.cph_selected), martingale_residuals), col = "red")
```

#### Observations:

-   A and B used as numeric covariates in the model (although tested at discrete levels) and C used as categorical covariate.
-   Martingale residuals do not show any clear pattern when plotted against covariates.
-   Martingale residuals do not show any clear pattern when plotted against fitted values
-   Functional form of the covariates is correctly specified in the model.

### Checking CPH assumption: Deletion residuals (DFBETAS)

```{r}
# Residuals with deletion
deletion_residuals <- residuals(fit.cph_selected, type = "dfbetas")
deletion_residuals_df <- sqrt(rowSums(deletion_residuals^2)) # Calculate the overall influence of each observation
df$Resid.DFBETAS <- deletion_residuals_df

# Plot the deletion residuals
plot(df$Resid.DFBETAS, type = "h", 
     xlab = "Observation Index", 
     ylab = "DFBETAS Residuals", 
     main = "DFBETAS Residuals for CPH Model")
abline(h = 0, col = "red")
```

#### Observations:

-   No observation has a DFBETAS residual that stands out as an influential outlier.

### Checking CPH assumption: Proportionality of hazards (log-log plots)

```{r}
# Proportionality of hazards log-log plots

surv = ddply(df, 
             factor_names, 
             function(subdf) {
               fit <- survfit(Surv(Cycles, Censoring) ~ 1, data = subdf)
               data.frame(
                 time = fit$time, 
                 surv = fit$surv, 
                 group = paste(subdf[1, factor_names], collapse = "_"))
               }
             )

ggplot(surv, aes(x = time, y = -log(-log(surv)), color = group)) +
  geom_step() +
  coord_trans(x = "log", y = "identity") +
  labs(title = "Log-Log Survival Curves for Proportional Hazards Check",
       x = "Time (Cycles)",
       y = "-log(-log(Survival Probability))")

```

#### Observations:

-   Log-log plots show approximately parallel curves for different groups, suggesting that the proportional hazards assumption is reasonable for the selected CPH model.

### Checking CPH assumption: Schoenfeld residuals

```{r}
# Check proportional hazards assumption
residual.sch <- cox.zph(fit.cph_selected)
print(residual.sch)

# Residuals vs. Time (Cycles) for each covariate
plot(residual.sch, var = "A.Stroke", main = "Schoenfeld Residuals for A.Stroke")
plot(residual.sch, var = "B.ProcessTpt", main = "Schoenfeld Residuals for B.ProcessTpt")
plot(residual.sch, var = "C.Method", main = "Schoenfeld Residuals for C.Method")
```

#### Observations:

-   The global test for proportional hazards assumption is not significant (p-values \> 0.05), suggesting that the covariate effects are constant over time.
-   Smoothed Schoenfeld residual plots vs. Time (Cycles) shows an approximately flat line, further supporting the proportional hazards assumption.
